sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-27 doc/paper.Rmd
Sections; lme4
Add missing E from Fig 2

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to spatial models that use the SPDE, but either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some of the flexibility for the particular class of models discussed here.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges. VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data. inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here. It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021]. We provide a more detailed comparison of related software packages in Table 1 and the discussion. -->

Here, we introduce our R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace any of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface for a specific class of spatial and spatiotemporal models that should be familiar to users of lme4, glmmTMB, or mgcv [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of related software.

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(a) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
The circle area corresponds to density of Pacific Cod for individual trawls.
(b) Conditional effect of depth modelled as a penalized smoother.
The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
These predictions are made omitting the spatial random field.
(c) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(d) Overall prediction: these estimates represent the combination of all fixed and random effects.

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts (**REF**) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):

@@

## Model selection and validation

@@

## Parameter priors/penalties

@@

## Derived quantities of interest

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is a less modular interface (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.

>>>>

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
This residual correlation is ideally dealt with because doing so (1) allows for valid statistical inference [@legendre1989a], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d, @barnett2021].
<!-- I'm finding the above a bit awkward. What about: Ideally, this residual correlation should be accounted for because doing so... OR Accounting for this residual correlation is important because doing so... -->

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
Similarly to how random intercepts can account for correlation among groups (e.g., plants counted in plots of land), spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
<!-- the e.g. here isn't intuitive for me... usually you only get one count from one plot so what is being grouped? Maybe heights of trees on different plots of land? -->
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or some of the flexibility for the particular class of models discussed here.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges. VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data. inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here. It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021]. We provide a more detailed comparison of related software packages in Table 1 and the discussion. -->

Here, we introduce our R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv, for a specific class of spatial and spatiotemporal models [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through TODO case studies, and concludes with a discussion of related software.

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
The circle area corresponds to density of Pacific Cod for individual trawls.
The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
These predictions are made omitting the spatial random field.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
(D) Conditional effect of depth modelled as a penalized smoother.
(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (**see appendix for more detail**):

@@

# Model selection and validation

@@

# Parameter priors/penalties

@@

# Derived quantities of interest

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is less modular (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
